Do researchers preferentially collaborate with same-gendered colleagues?

### 2018-04-27 manuscript/Manuscript.Rmd
More writing

---
title: "Do researchers preferentially collaborate with same-gendered colleagues?"
author:
- Luke Holman* and Claire Morandin
- '*\textit{luke.holman@unimelb.edu.au} \vspace{5mm}'
output:
  pdf_document: default
  html_document: default
bibliography: references.bib
fontsize: 12pt
header-includes: 
  \usepackage[all]{hypcap} 
  \usepackage{amsmath} 
  \usepackage{booktabs}
  \usepackage{caption} 
  \usepackage[labelfont=bf]{caption} 
  \usepackage{titling} 
  \pretitle{\begin{flushleft}\LARGE}
  \posttitle{\par\end{flushleft}\vskip 0.5em} 
  \preauthor{\begin{flushleft}\large \lineskip 0.5em} 
  \postauthor{\par\end{flushleft}} 
  \predate{\begin{flushleft}\large} 
  \postdate{\par\end{flushleft}}
  \usepackage{fancyhdr} 
  \pagestyle{fancy} 
  \fancyhead[LO,LE]{\textsl{\leftmark}} 
  \rhead[]{Conference gender ratios} 
  \usepackage[noabbrev,capitalise]{cleveref}
  \usepackage{titlefoot} 
  \usepackage{amssymb} 
  \usepackage{rotating}
csl: the-american-naturalist.csl
abstract: Evidence suggests that women in academia are hindered by conscious and unconscious biases, and that many female researchers feel excluded from formal and informal opportunities for research collaboration. In addition to helping redress gender imbalance in the academic workforce, increasing women’s access to collaboration helps scientific progress by drawing on more of the available human capital. Here, we test whether researchers preferentially collaborate with same-gendered colleagues, expanding on earlier work by using more stringent methods and a larger dataset. We reaffirm that researchers tend to co-publish with colleagues of the same gender, and show that this ‘gender homophily’ is slightly stronger today than it was 10 years ago. Contrary to our expectations, we found no evidence that homophily is driven mostly by senior academics, and no evidence that homophily is strongest in fields where women are in the minority. Interestingly, homophily was negatively correlated with journal impact factor (standardised by research discipline), as predicted if mixed-gender teams produce better research. \vspace{5mm}
  \par\noindent \textbf{Keywords:} Gender bias, Homophily, Scientific collaboration, Test mining, Women in STEM
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

@@

# Methods
## Obtaining author lists and assigning gender
We used the dataset collected from PubMed by Holman et al. [-@Holman_2018]. Briefly, Holman et al. created the dataset by downloading every single article indexed on PubMed, from the oldest articles up to the present (which was August 2016), and attempting to assign a gender to each author based on their given name, using the web service genderize.io. Each journal was assigned to one of >100 scientific disciplines (e.g. “Nephrology” or “Cell Biology”), using PubMed’s journal categorisations where available, or manual methods otherwise. Because the present study focuses on co-authorship, all single-author papers were removed from the dataset. To simplify the statistical analysis, we also discarded all papers for which we could not determine the gender of every author with at least 95% certainty. To mitigate Wahlund effects caused by variation in the gender ratio of researchers over time (see below), we also discarded all papers except those that were published 0-1 or 10-11 years ago, as of August 2016. Lastly, we excluded journals with fewer than 50 suitable papers. This left us with a dataset from August 2015 - August 2016 containing 276,879 papers with 1,311,213 authorships, and a dataset from August 2005 - August 2006 containing 151,652 papers and 647,634 authorships. These papers came from 2077 and 1192 journals respectively, which were grouped into 107 and 101 research disciplines. There was a median of 88 (87) papers per journal, 419 (371) authors per journal, and 4 (4) authors per paper (the first number is for the 2015-6 dataset, and the bracketed number for the 2005-6 dataset).

## Calculating $\alpha$, the coefficient of homophily
Inspired by Bergstrom et al. (http://www.eigenfactor.org/gender/assortativity/measuring_homophily.pdf), we defined the coefficient of homophily as $\alpha = p - q$, where $p$ is the probability that a randomly-chosen co-author of a male author is a man, and $q$ is the probability that a randomly-chosen co-author of a female author is a man. Therefore, $\alpha$ > 0 suggests that same-gender authors publish together more often than expected if authors assort randomly with respect to gender (homophily), while $\alpha$ < 0 suggests that opposite-gender authors publish together more often than expected (heterophily), and $\alpha$ = 0 suggests random assortment with respect to gender.

@@

Our test assumes that the expected value of $\alpha$ is zero if authors randomly assort, but for small datasets this is not always true (as pointed out by Carl T. Bergstrom in a blog post, http://www.eigenfactor.org/gender/assortativity/note_to_eisen.rtf). To borrow Prof. Bergstrom’s example, consider a small research specialty comprising just two men and two women researchers, who have together produced six two-author papers: one in each of the six possible combinations. For these six papers, $\alpha$ = $-\frac{1}{3}$, even though same- and opposite-gendered coauthors were selected in equal proportion to their frequency in the pool of possible collaborators. To control for the fact that the null expectation for $\alpha$ is not zero for every dataset (particularly small datasets), we devised an adjusted version of the coefficient of homophily, which we term $\alpha'$. Every time we calculated $\alpha$ for a set of papers, we also determined the expected value of $\alpha$ under the null hypothesis that authors assort randomly with respect to gender. This was accomplished by randomly shuffling authors across papers 1000 times, recalculating $\alpha$, and taking the median. We then calculated $\alpha'$ by subtracting the null expectation for $\alpha$ from the observed value. We also used the null-simulated $\alpha$ values to calculate a two-tailed p-value for the observed value of $\alpha$; the p-value was defined as the proportion of null simulations for which $|\alpha_{null}| > |\alpha_{obs}|$. We applied false discovery rate correction to each set of p-values to account for multiple testing (Benjamini and Hochberg 1995).

@@

## Minimising the Wahlund effect: discipline and time period
As discussed in the Introduction, the Wahlund effect can give the appearance of gender homophily (reflected in our study as $\alpha'$ > 0) whenever one has lumped together different ‘subpopulations’ of literature that differ in author gender ratio (Figure 1).

To minimise bias in $\alpha'$ due to the Wahlund effect, we restricted each set of papers to a single research specialty to the greatest extent allowed by our data. We therefore decided to only calculate $\alpha'$ for individual journals, since papers from the same journal typically focus on one or a few closely related topics. Although some journals, e.g. PLoS ONE, publish research from diverse disciplines with very different author gender ratios [@Holman_2018], calculating $\alpha'$ for these highly multidisciplinary journals is still useful as a contrast. The difference in $\alpha'$ between highly multidisciplinary and more specialised journals, e.g. PLoS ONE versus PLoS Computational Biology, gives a rough estimate of the maximum extent to which multidisciplinarity can inflate $\alpha'$.

As well as varying between disciplines, the gender ratio of authors has changed markedly over time [@Holman_2018]. Because the gender ratio was more male-biased in the past, $\alpha'$ would be inflated if we calculated it for a sample of papers published over a lengthy time frame. To minimise this effect, we only sampled papers from two one-year periods (namely 2005-6 and 2015-16). The median change per year in % (fe)male authors across journals is < 0.5% [@Holman_2018], and so restricting our dataset to a single year should prevent temporal changes in gender ratio from noticeably inflating our estimates of $\alpha'$.

@@

To address this, we also analysed every combination of journal and author country of affiliation for which we had enough data (meaning 50 or more papers published in 2015-16). For simplicity, we restricted the dataset to only include papers for which Holman et al. [-@Holman_2018] had successfully identified the country of affiliation for all authors on the paper, and all authors shared the same country of affiliation. Restricting the dataset in this fashion produced enough data to measure $\alpha'$ for 325 combinations of journal and country (median: 70 papers and 273 authors per combination).

## Calculating standardised journal impact factor
We obtained the 3-year impact factor for each journal from Clarivate Analytics. To account for large differences in impact factor between disciplines, we took the the residuals from a model with log10 impact factor as the response and the research discipline of the journal as a random effect. Thus, journals with a positive standardised impact factor have a higher mean number of citations than the average for journals in their discipline. We then tested whether our estimates of gender homophily covary with journal impact factor using Spearman rank correlation.

## Statistical models
Previous authors (e.g. Bonham & Stefan 2017) have hypothesised that senior scientists preferentially recruit staff and students of the same gender, and/or that junior researchers preferentially select same-gendered mentors. In the majority of disciplines, authorship conventions mean that the first-listed author is often an early-career researcher, while the author listed last is more likely to be a senior researcher leading a research team (Wren et al. 2007). Assuming that senior researchers are the main drivers of homophily and that there are enough papers with three or more authors, we predict that the last author’s gender will be the strongest predictor of the remaining authors’ genders (i.e. the gender of the last author will be more salient than that of the first author, or any other authorship position). This is because the first author’s gender would simply be an imperfect correlate of the true causal effect, while the last author’s gender would be the causal effect itself.

To test whether $\alpha'$ for last authors tends to be higher than $\alpha'$ for first authors for any given dataset, we used a linear mixed model implemented in the ``lme4`` and ``lmerTest`` packages for R, with _authorship position_ (first or last) as a fixed factor, and _journal_ and _research discipline_ as crossed random effects. The response variable was $\alpha'$, and we weighted each observation by the inverse of the standard error from our estimate of $\alpha'$, meaning that more accurate measurements of $\alpha'$ had more influence on the results.

We used a very similar model to test for a difference in $\alpha'$ between the 2005-6 and the 2015-16 datasets. There were two differences: instead of authorship position, we fit year range as a two-level fixed factor, and used $\alpha'$ estimated for all authors (not first/last authors) as the response variable.

@@

# Results
## Most disciplines show evidence of gender homophily
Figure 2 shows the distribution of $\alpha'$ estimates in 2015-2016 across all journals for which we recovered sufficient data, when $\alpha'$ was calculated for all authors, first authors only, or last authors only. The great majority of journals had $\alpha'$ > 0, and for many of these the corrected two-tailed p-values suggested that $\alpha'$ was significantly greater than zero (1469/2077 journals were significant in 2015-16, and 404/1192 in 2005-6). Only 2/2077 journals had statistically significantly heterophily (i.e. $\alpha'$ < 0) in 2015-16, and 1/1192 in 2005-6. The remaining 606 or 787 journals (in 2015 and 2005 respectively) had a value of $\alpha'$ not significantly different from zero, such that we could not reject the null hypothesis of random assortment with respect to gender.

$\alpha'$ was statistically significantly higher in the sample from 2015-16 relative to 2005-6, but the difference in means was small (Figure SX; Effect of the fixed factor ‘time period’ in a linear mixed model of the data for all author positions: Cohen’s $d$ = 0.09  0.04, $t_{948}$ = 2.51, p = 0.012). This means that gender homophily was slightly stronger in 2015-6 relative to 2005-6.
When comparing pairs of $\alpha'$ values estimated for the first and last authors for the same journals, we found that $\alpha'$ tended to be higher for the first authors than for the last authors (Figure SX; Effect of the fixed factor ‘Authorship position’ in a linear mixed model: Cohen’s d = 0.07 ± 0.02, $t_{1988}$ = 4.48, p < 0.0001). This suggests that the gender of the first author was a slightly stronger predictor of the remaining authors’ genders than the gender of the last author, i.e. the opposite of our prediction based on the hypothesis that senior scientists are causally response for homophily.

## Possible differences in the strength of homophily between disciplines
Figure 3 illustrates the variance in journal homophily values ($\alpha'$) across scientific disciplines. All disciplines had a positive average $\alpha'$, although homophily appeared somewhat stronger in some disciplines than others (e.g. mean $\alpha'$ was 0.12±0.02 for Urology journals, and 0.03±0.01 for Veterinary Medicine journals; Table S3). However, there was little evidence for strong differences in $\alpha'$ between disciplines: the random factor ‘discipline’ explained at most 1% of the variance in $\alpha'$ in the two linear mixed models described in the previous section (see mixed models in Online Supplementary Material). This implies that the processes causing positive $\alpha'$ are similarly strong in all the disciplines we examined.

Reassuringly, we found no support for our prediction that journals publishing on a wide range of topics should have high $\alpha'$ values due to the Wahlund effect (Figure 3). For example, the journal category “Multidisciplinary” - which includes journals like PLoS ONE, Nature, Science, and PNAS - does not have notably elevated $\alpha'$. This result suggests that our estimates of homophily were not greatly inflated by the presence of disparate research topics (with variable author gender ratios) in the same journal.

## Relationship between gender homophily and gender ratio
One might predict that researchers would be more likely to seek out same-gendered colleagues in the most strongly gender-biased disciplines in our dataset (e.g. Surgery and Nursing), relative to disciplines with a comparatively gender-balanced workforce (e.g. Psychiatry). However, we instead found evidence of the reverse: there was a positive, non-linear relationship between the overall gender ratio of all authors publishing in the journal  [as estimated by @Holman_2018], and the estimated value of $\alpha'$ for all authors and for first authors (Figure 4). That is, journals with a balanced or female-biased author gender ratio tended to have higher $\alpha'$ than journals with a male-biased author gender ratio (GAM smooth terms p < 0.001, see Online Supplementary Material). The relationship was not statistically significant when $\alpha'$ was calculated for last authors (GAM, p = 0.142), though the trend appeared similar (Figure 4).

## Relationship between journal impact factor and gender homophily
We observed a noisy but statistically significant linear relationship between discipline-standardised journal impact factor and $\alpha'$, such that journals with a high impact factor for their discipline had weaker gender homophily (Figure 5; linear regression: $R^2$ = 0.043, $t_{1415}$ = -8.0, p < 0.0001).

@@

# Discussion
We found evidence that researchers tend to publish with same-gendered coauthors, even after implementing stringent controls for ‘spurious’ homophily resulting from Wahlund effects (\autoref{wahlund_plot}). Our data therefore concur with previous conclusions (Ferber and Teiman 1980; McDowell and Smith 1992; Crow and Smykla 2015; Ghiasi et al. 2015; Zettler et al. 2016; Fahmy and Young 2016; Jadihi et al. 2017; Araujo and Fontainha 2017a; Teele and Thelen 2017) that research collaborations occur between same-gendered colleagues substantially more often than expected if collaborators are selected without regard to their gender. Moreover, we found that gender homophily is not unique to particular disciplines, but rather appears to be similarly strong across many diverse fields of STEMM.

@@

\newpage
# Figures
![The Wahlund effect can make it appear that authors prefer to publish with same-gendered colleagues, even when no such preference exists. Here, the circles represent male (blue) and female (orange) authors, and coauthors are linked with lines. Across the whole set of ten papers, there is an apparent excess of same-gender collaborations. Specifically, there are six same-gender papers and only four mixed-gender papers, which is less than the $10\times2\times0.5\times0.5 = 5$ mixed-gender papers we would expect under the null hypothesis that authors assort randomly with respect to gender. However, within each subset, there is no evidence that authors prefer to publish with same-gendered individuals. The Wahlund effect will tend to inflate the frequency of same-sex coauthors whenever the data is composed of two or more disconnected subsets of literature with different author gender ratios; these subsets could be research discipline, old versus new papers, or papers from authors in different countries. \label{wahlund_plot}](../figures/Figure 1 - Wahlund figure.png)

\newpage
```{r echo=FALSE, results='hide'}
fig2 <- readRDS("fig2_numbers.RData")
n.journals <- fig2[1]
n.sig.homo <- fig2[2]
n.sig.hetero <- fig2[3]
```
![Of the `r n.journals` journals for which we had adequate data in 2015-2016, `r n.sig.homo` showed statistically significant evidence of homophily (denoted by $\alpha' > 0$), and `r n.sig.hetero` showed statistically significant evidence of heterophily ($\alpha' < 0$), after adjusting p-values using Benjamini-Hochberg false discovery rate correction. The white area shows the number of journals for which homophily was significantly stronger than expected under the null hypothesis (p < 0.05), while the blue area shows all the remainder. Patterns were similar whether $\alpha'$ was calculated for all authors, for first authors only, or for last authors only. \label{alpha_histograms}](../figures/figure2.pdf)

\newpage
![There is a weakly positive, non-linear relationship between the gender ratio of authors publishing in a journal, and the coefficient of homophily ($\alpha'$). Specifically, journals with 50% women authors or higher tended to have more same-sex coauthorships than did journals with predominantly men authors.  This relationship held whether $\alpha'$ was calculated for all authors, first authors only, or last authors only. A negative value on the x axis denotes an excess of men authors, a positive value denotes an excess of women authors, and zero denotes gender parity. The lines were fitted using generalised additive models with the smoothing parameter $k$ set to 3. \label{alpha_gender_ratio}](../figures/figure3.pdf)

\newpage
![Histogram of $\alpha'$ for 325 unique combinations of journal and country, using data from August 2015 - August 2016. The white areas denote combinations for which $\alpha'$ differs significantly from zero (p < 0.05, following false discovery rate correction). \label{alpha_country}](../figures/figure4.pdf)

<!-- Sensitivity analysis of A -->
<!-- Note: using Latex to avoid an issue where the heading appears below the first figure. -->

>>>>

---
title: "Do researchers preferentially collaborate with same-gendered colleagues?"
author:
- Luke Holman* and Claire Morandin
- '*\textit{luke.holman@unimelb.edu.au} \vspace{5mm}'
output:
  pdf_document: default
  html_document: default
bibliography: references.bib
fontsize: 12pt
header-includes: 
  \usepackage[all]{hypcap} 
  \usepackage{amsmath} 
  \usepackage{booktabs}
  \usepackage{caption} 
  \usepackage[labelfont=bf]{caption} 
  \usepackage{titling} 
  \pretitle{\begin{flushleft}\LARGE}
  \posttitle{\par\end{flushleft}\vskip 0.5em} 
  \preauthor{\begin{flushleft}\large \lineskip 0.5em} 
  \postauthor{\par\end{flushleft}} 
  \predate{\begin{flushleft}\large} 
  \postdate{\par\end{flushleft}}
  \usepackage{fancyhdr} 
  \pagestyle{fancy} 
  \fancyhead[LO,LE]{\textsl{\leftmark}} 
  \rhead[]{Conference gender ratios} 
  \usepackage[noabbrev,capitalise]{cleveref}
  \usepackage{titlefoot} 
  \usepackage{amssymb} 
  \usepackage{rotating}
csl: the-american-naturalist.csl
abstract: Evidence suggests that women in academia are hindered by conscious and unconscious biases, and that many female researchers feel excluded from formal and informal opportunities for research collaboration. In addition to ensuring fairness and helping redress gender imbalance in the academic workforce, increasing women’s access to collaboration could help scientific progress by drawing on more of the available human capital. Here, we test whether researchers preferentially collaborate with same-gendered colleagues, using more stringent methods and a larger dataset than in past work. Our results reaffirm that researchers co-publish with colleagues of the same gender, and show that this ‘gender homophily’ is slightly stronger today than it was 10 years ago. Contrary to our expectations, we found no evidence that homophily is driven mostly by senior academics, and no evidence that homophily is strongest in fields where women are in the minority. Interestingly, homophily was negatively correlated with journal impact factor (standardised by research discipline), as predicted if mixed-gender teams produce better research. \vspace{5mm} \par\noindent \textbf{Keywords:} Gender bias, Homophily, Scientific collaboration, Text mining, Women in STEM.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Load some little files with numbers to be inserted into the MS
sample.size <- read.csv("sample.size.table.csv", stringsAsFactors = FALSE)
sample.size[,2] <- prettyNum(sample.size[,2], big.mark = ",")
sample.size[,3] <- prettyNum(sample.size[,3], big.mark = ",")
names(sample.size)[2:3] <- c("n2005", "n2015")

fig2 <- readRDS("fig2_numbers.RData") # For figure 2
n.journals <- fig2[1]
n.sig.homo <- fig2[2]
n.sig.hetero <- fig2[3]

R2 <- readRDS("R2_for_figure4.RData") # For Fig 4
```

@@

# Methods
## Obtaining author lists and assigning gender
We used the dataset collected from PubMed by Holman et al. [-@Holman_2018]. Briefly, that dataset was created by downloading every single article indexed on PubMed, from the oldest articles up to the present (which was August 2016), and attempting to assign a gender to each author based on their given name, using the web service genderize.io. Each journal was assigned to one of 107 scientific disciplines (e.g. “Nephrology” or “Cell Biology”), using PubMed’s journal categorisations where available. Because the present study focuses on co-authorship, all single-author papers were removed from the dataset. To simplify the statistical analysis, we also discarded all papers for which we could not determine the gender of every author with at least 95% certainty. To mitigate Wahlund effects caused by variation in the gender ratio of researchers over time (see below), we also discarded all papers except those that were published 0-1 or 10-11 years before the PubMed data were collected (i.e. 20 August 2016). Lastly, we excluded journals with fewer than 50 suitable papers. This left us with a dataset from August 2015 - August 2016 containing `r sample.size$n2015[3]` papers with `r sample.size$n2015[4]` authorships, and a dataset from August 2005 - August 2006 containing `r sample.size$n2005[3]` papers and `r sample.size$n2005[4]` authorships. These papers came from `r sample.size$n2015[2]` and `r sample.size$n2005[2]` journals respectively, which were grouped into `r sample.size$n2015[1]` and `r sample.size$n2005[1]` research disciplines. There was a median of `r sample.size$n2015[5]` (`r sample.size$n2005[5]`) papers per journal, `r sample.size$n2015[6]` (`r sample.size$n2005[6]`) authors per journal, and `r sample.size$n2015[7]` (`r sample.size$n2005[7]`) authors per paper (the first number is for 2015-6, and the bracketed number for 2005-6).

## Calculating $\alpha$, the coefficient of homophily
Inspired by Bergstrom et al. (http://www.eigenfactor.org/gender/assortativity/measuring_homophily.pdf), we defined the coefficient of homophily as $\alpha = p - q$, where $p$ is the probability that a randomly-chosen co-author of a _male_ author is a man and $q$ is the probability that a randomly-chosen co-author of a _female_ author is a man. Therefore, $\alpha$ > 0 suggests that same-gender authors publish together more often than expected if authors assort randomly with respect to gender (homophily), $\alpha$ < 0 suggests that opposite-gender authors publish together more often than expected (heterophily), and $\alpha$ = 0 suggests random assortment with respect to gender.

@@

Our test assumes that the expected value of $\alpha$ is zero if authors randomly assort, but for small datasets this assumption is not always true (as pointed out by Carl T. Bergstrom in a blog post, http://www.eigenfactor.org/gender/assortativity/note_to_eisen.rtf). To borrow Prof. Bergstrom’s example, consider a small research specialty comprising just two men and two women researchers, who have together produced six two-author papers: one paper in each of the six possible two-author combinations. For these six papers, $\alpha$ = $-\frac{1}{3}$, even though same- and opposite-gendered coauthors were selected in equal proportion to their frequency in the pool of possible collaborators. To control for the fact that the null expectation for $\alpha$ is not necessarily zero (particularly for very small datasets), we devised an adjusted version of the coefficient of homophily, which we term $\alpha'$. Every time we calculated $\alpha$ for a set of papers, we also determined the expected value of $\alpha$ under the null hypothesis that authors assort randomly with respect to gender. This was accomplished by randomly shuffling authors across papers 1000 times, recalculating $\alpha$, and taking the median. We then calculated $\alpha'$ by subtracting the null expectation for $\alpha$ from the observed value. We also used the null-simulated $\alpha$ values to calculate a two-tailed p-value for the observed value of $\alpha$; the p-value was defined as the proportion of null simulations for which $|\alpha_{null}| > |\alpha_{obs}|$. We applied false discovery rate correction to each set of p-values to account for multiple testing (Benjamini and Hochberg 1995).

@@

## Minimising the Wahlund effect: discipline and time period
As discussed in the Introduction, the Wahlund effect can give the appearance of gender homophily (reflected in our study as $\alpha'$ > 0) whenever one lumps together different ‘subpopulations’ of academic papers that differ in their author gender ratios (Figure 1).

To minimise bias in $\alpha'$ due to the Wahlund effect, we restricted each set of papers to a single research specialty to the greatest extent allowed by our data. Specifically, we only calculated $\alpha'$ for individual journals, since papers from the same journal typically focus on one or a few closely related topics. Although some journals, e.g. _PLoS ONE_, publish research from diverse disciplines with very different author gender ratios [@Holman_2018], calculating $\alpha'$ for these highly multidisciplinary journals is still useful as a contrast. The difference in $\alpha'$ between highly multidisciplinary and more specialised journals, e.g. _PLoS ONE_ versus _PLoS Computational Biology_, gives a rough estimate of the extent to which multidisciplinarity can inflate $\alpha'$.

As well as varying between disciplines, the gender ratio of authors has changed markedly over time [@Holman_2018]. Because the gender ratio was more male-biased in the past, $\alpha'$ would be inflated if we calculated it for a sample of papers published over a long time frame. To minimise this effect, we only sampled papers from two one-year periods (namely 2005-6 and 2015-16). The median change per year in % (fe)male authors across journals is < 0.5% [@Holman_2018], and so restricting our dataset to a single year should prevent temporal changes in gender ratio from noticeably inflating our estimates of $\alpha'$.

@@

To address this, we also analysed every combination of journal and author country of affiliation for which we had enough data (i.e. 50 or more papers published in 2015-16). For simplicity, we restricted the dataset to only include papers for which Holman et al. [-@Holman_2018] had successfully identified the country of affiliation for all authors on the paper, and all authors shared the same country of affiliation. Restricting the dataset in this fashion produced enough data to measure $\alpha'$ for 325 combinations of journal and country (median: 70 papers and 273 authors per combination).

## Calculating standardised journal impact factor
We obtained the 3-year impact factor for each journal from Clarivate Analytics. To account for large differences in impact factor between disciplines, we took the the residuals from a model with $Log_{10}$ impact factor as the response and the research discipline of the journal as a random effect. Thus, journals with a positive standardised impact factor have a higher mean number of citations than the average for journals in their discipline. We then used Spearman rank correlation to test whether our estimates of gender homophily covaried with journal impact factor.

## Statistical analysis
Previous authors (e.g. Bonham & Stefan 2017) have hypothesised that senior scientists preferentially recruit staff and students of the same gender, and/or that junior researchers preferentially select same-gendered mentors. In the majority of disciplines, authorship conventions mean that the first-listed author is often an early-career researcher, while the author listed last is more likely to be a senior researcher leading a research team (Wren et al. 2007). Assuming that senior researchers are the main drivers of homophily and that there are enough papers with three or more authors, we predict that the last author’s gender will be the strongest predictor of the remaining authors’ genders (i.e. the gender of the last author will be more salient than that of the first author, or any other authorship position). This is because the first author’s gender would simply be an imperfect correlate of the true causal effect, while the last author’s gender would be the causal effect itself.

To test whether $\alpha'$ for last authors tends to be higher than $\alpha'$ for first authors for any given dataset, we used a linear mixed model implemented in the ``lme4`` and ``lmerTest`` packages for R, with _authorship position_ (first or last) as a fixed factor, and _journal_ and _research discipline_ as crossed random effects. The response variable was $\alpha'$, and we weighted each observation by the inverse of the standard error from our estimate of $\alpha'$, meaning that more accurate measurements of $\alpha'$ had more influence on the results. We used a similar model to test for a difference in $\alpha'$ between the 2005-6 and the 2015-16 datasets, with two differences: instead of authorship position, we fit year range as a two-level fixed factor, and used $\alpha'$ estimated for all authors (not first/last authors) as the response variable.

@@

# Results
## Gender homophily by discipline, time period, and authorship position
\autoref{alpha_histograms} shows the distribution of $\alpha'$ estimates in 2015-2016 across all journals for which we recovered sufficient data, when $\alpha'$ was calculated for all authors, first authors only, or last authors only. The great majority of journals had $\alpha'$ > 0, and for many of these the corrected two-tailed p-values suggested that $\alpha'$ was significantly greater than zero (1469/2077 journals were significant in 2015-16, and 404/1192 in 2005-6). Only 2/2077 journals had statistically significantly heterophily (i.e. $\alpha'$ < 0) in 2015-16, and 1/1192 in 2005-6. The remaining 606 or 787 journals (in 2015 and 2005 respectively) had a value of $\alpha'$ not significantly different from zero, such that we could not reject the null hypothesis of random assortment with respect to gender.

$\alpha'$ was significantly higher in the sample from 2015-16 relative to 2005-6, but the difference in means was small (Figure S3; Effect of the fixed factor ‘time period’ in a linear mixed model of the data for all author positions: Cohen’s $d$ = 0.09 ± 0.04, $t_{948}$ = 2.51, p = 0.012). This result illustrates that gender homophily was slightly stronger in 2015-6 relative to 2005-6.

When comparing pairs of $\alpha'$ values estimated for the first and last authors for the same journals, we found that $\alpha'$ tended to be higher for first authors than for last authors (Figure S4; Effect of the fixed factor ‘Authorship position’ in a linear mixed model: Cohen’s d = 0.07 ± 0.02, $t_{1988}$ = 4.48, p < 0.0001). This suggests that the gender of the first author was a slightly stronger predictor of the remaining authors’ genders than the gender of the last author, i.e. the opposite of our prediction based on the hypothesis that senior scientists are causally responsible for homophily.

## Possible differences in the strength of homophily between disciplines
\label{alpha_gender_ratio} illustrates the variance in journal homophily values ($\alpha'$) across scientific disciplines. All disciplines had a positive average $\alpha'$, although homophily appeared somewhat stronger in some disciplines than others (e.g. mean $\alpha'$ was 0.12 ± 0.02 for Urology journals, and 0.03 ± 0.01 for Veterinary Medicine journals; Table S3). However, there was little evidence for strong differences in $\alpha'$ between disciplines: the random factor ‘discipline’ explained at most 1% of the variance in $\alpha'$ in the two linear mixed models described in the previous section (see mixed models in Online Supplementary Material). This implies that the processes causing positive $\alpha'$ are similarly strong in all the disciplines we examined.

Reassuringly, we found no support for our prediction that journals publishing on a wide range of topics should have high $\alpha'$ values due to the Wahlund effect \label{alpha_gender_ratio}. For example, the journal category “Multidisciplinary” - which includes journals like PLoS ONE, Nature, Science, and PNAS - does not have notably elevated $\alpha'$. This result suggests that our estimates of homophily were not greatly inflated by the presence of disparate research topics (with variable author gender ratios) in the same journal.

## Relationship between gender homophily and gender ratio
One might predict that researchers would be more likely to seek out same-gendered colleagues in strongly gender-biased disciplines (e.g. Surgery or Nursing), relative to disciplines with a comparatively gender-balanced workforce (e.g. Psychiatry). However, we instead found evidence of the reverse: there was a positive, non-linear relationship between the overall gender ratio of all authors publishing in a particular journal [as estimated by @Holman_2018], and the estimated value of $\alpha'$ for all authors and for first authors \label{alpha_country}. That is, journals with a balanced or female-biased author gender ratio tended to have higher $\alpha'$ than journals with a male-biased author gender ratio (GAM smooth terms p < 0.001, see Online Supplementary Material). The relationship was not statistically significant when $\alpha'$ was calculated for last authors (GAM, p = 0.142), though the trend appeared similar \label{alpha_country}.

## Relationship between journal impact factor and gender homophily
We observed a noisy but statistically significant linear relationship between discipline-standardised journal impact factor and $\alpha'$, such that journals with a high impact factor for their discipline had weaker gender homophily than did journals with a low impact factor for their discipline (Figure 5; linear regression: $R^2$ = `r R2`, $t_{1415}$ = -8.0, p < 0.0001).

@@

# Discussion
We found evidence that researchers tend to publish with same-gendered coauthors, even after implementing stringent controls for ‘spurious’ homophily resulting from Wahlund effects (\autoref{wahlund_plot}). Our data therefore reaffirm earlier conclusions (Ferber and Teiman 1980; McDowell and Smith 1992; Crow and Smykla 2015; Ghiasi et al. 2015; Zettler et al. 2016; Fahmy and Young 2016; Jadihi et al. 2017; Araujo and Fontainha 2017a; Teele and Thelen 2017) that research collaborations occur between same-gendered colleagues substantially more often than expected if collaborators are selected without regard to their gender. Many journals had $\alpha'$ values over 0.1, meaning that the gender ratio of the average man's coauthors is skewed towards men by >10%, relative to the gender ratio of a woman's coauthors (or equivalently, that the gender ratio of women's coauthors is >10% more female than that of men's). Few journals had $\alpha'$ values below zero, and fewer still had significantly negative $\alpha'$ values, suggesting that researchers rarely seek out collaborators of the opposite gender (or at least, that they do so less often than they seek out same-gendered colleagues).

We found no evidence that gender homophily is restricted to particular disciplines: instead, there was evidence of homophily across many diverse research disciplines. We also found that gender homophily was marginally stronger in 2015-2016, relative to 2005-2006. This result suggests that recent increases in the number of women working in STEMM, as well as recent efforts to recruit, retain, and celebrate women in STEMM, have (if anything) strengthened gender homophily rather than weakening it.

Gender homophily was slightly stronger in research disciplines with a female-biased gender ratio, as compared to male-biased or gender-balanced fields. One interpretation of this result is that men are more likely to seek out other men in fields in which men are in the minority, relative to the homophily shown by women in fields in which women are in the minority.

We also found that journals with a high impact factor (relative to other journals in the same research discipline) tended to have weaker gender homophily than did low-impact journals. If one accepts the contentious claim that papers published in high-impact journals tend to be of higher quality than those in low-impact journals [e.g. @Garfield_2006], then this result provides correlational support for the hypothesis that mixed-gender teams produce better research than single-gender teams (Campbell et al. 2013).

Contrary to our expectations, we found no evidence that homophily is driven mostly by senior academics, and no evidence that homophily is strongest in fields where women are in the minority. Interestingly, homophily was negatively correlated with journal impact factor (standardised by research discipline), as predicted if mixed-gender teams produce better research.

@@

\newpage
# Figures
![The Wahlund effect can make it appear as if authors prefer to publish with same-gendered colleagues, even if no such preference exists. Here, coloured circles represent male and female authors, and coauthors are linked with lines. Across the whole set of ten papers, there is an apparent excess of same-gender collaborations. Specifically, there are six same-gender papers and only four mixed-gender papers, which is fewer than the $10\times2\times0.5\times0.5 = 5$ mixed-gender papers we would expect under the null hypothesis that authors assort randomly with respect to gender. However, within each subset, there is no evidence that authors prefer to publish with same-gendered individuals. The Wahlund effect will tend to inflate the frequency of same-sex coauthors whenever the data is composed of two or more disconnected subsets of literature with different author gender ratios; these subsets could be research disciplines, older versus newer papers, or papers from authors in different countries. \label{wahlund_plot}](../figures/Figure 1 - Wahlund figure.pdf)

\newpage
![Of the `r n.journals` journals for which we had adequate data in 2015-2016, `r n.sig.homo` showed statistically significant evidence of homophily (denoted by $\alpha' > 0$), and `r n.sig.hetero` showed statistically significant evidence of heterophily ($\alpha' < 0$), after adjusting p-values using Benjamini-Hochberg false discovery rate correction. The white area shows the number of journals for which homophily was significantly stronger than expected under the null hypothesis (p < 0.05), while the blue area shows all the remainder. Patterns were similar whether $\alpha'$ was calculated for all authors, for first authors only, or for last authors only. \label{alpha_histograms}](../figures/figure2.pdf)

\newpage
![There is a weakly positive, non-linear relationship between the gender ratio of authors publishing in a journal, and the coefficient of homophily ($\alpha'$). Specifically, journals with 50% women authors or higher tended to have more same-sex coauthorships than did journals with predominantly men authors.  This relationship held whether $\alpha'$ was calculated for all authors, first authors only, or last authors only. A negative value on the x-axis denotes an excess of men authors, a positive value denotes an excess of women authors, and zero denotes gender parity. The lines were fitted using generalised additive models with the smoothing parameter $k$ set to 3. \label{alpha_gender_ratio}](../figures/figure3.pdf)

\newpage
![Journal impact factor (expressed relative to the average for the discipline) is negatively correlated with $\alpha'$. The relationship is noisy ($R^2$ = `r R2`), but the results suggest that journals with strong homophily tend to have lower impact factors than journals with weak homophily in the same discipline. \label{impact_factor}](../figures/figure5.pdf)
